A Bioconductor workflow for processing and analysing spatial proteomics data

### 2018-04-17 bioc-workflow-resubmit.Rnw
more changes to pre-processing section; should now also build on travis

To make use of the full functionality of the \Biocpkg{pRoloc} software
one needs to import their data into R and prepare them as an
\texttt{MSnSet}. The \texttt{MSnSet} is a dedicated data structure for
the efficient manipulation and processing of mass spectrometry and
proteomics data in R. Figure \ref{fig:msnset} illustrates a simplified view of the
\texttt{MSnSet} structure; there exists 3 key sub-parts (termed slots)
to such a data object: (1) the \texttt{exprs} (short for
\textit{expression} data) slot for storing the quantitation data, (2)
the \texttt{fData} slot (short for \textit{feature}-metadata) for
storing the feature meta-data, and finally (3) the \texttt{pData} slot
(short for \textit{pheno}-metadata, i.e. sample phenotypic data) for
storing the sample meta-data.

@@

Another slot of interest is \texttt{processingData}, that logs the
processing \texttt{MSnSet} objects undergo. The processing log can be
accessed with the \texttt{processingData} function and is displayed
under \textit{Processing information} in the textual object summary
when an \texttt{MSnSet}'s name is typed in the R console.

@@

There are a number of ways to import quantitation data and create an
\texttt{MSnSet} instance. All methods are described in the
\Biocpkg{MSnbase}
\href{http://bioconductor.org/packages/release/bioc/vignettes/MSnbase/inst/doc/MSnbase-io.pdf}{input/output
  capabilities vignette}. One suggested simple method is to use the
function \texttt{readMSnSet2}. The function takes a single spreadsheet
file name as input and extracts the columns containing the
quantitation data, as identified by the argument \texttt{ecol}, to
create the expression data, while the other columns in the spreadsheet
are appended to the feature meta-data slot.  By example, in the code
chunk below we read in the \texttt{csv} spreadsheet containing the
quantitation data from the intersect of replicates 1 and 2 of the
mouse map \cite{hyper}, using the \texttt{readMSnSet2} function. The
data is as available online with the manuscript (see tab 2 of the
\texttt{xlsx} supplementary data set 1 in \cite{hyper}, which should
be exported as a text-based spreadsheet). It is also available as a
\texttt{csv} in the Bioconductor \Biocexptpkg{pRolocdata} data
package, which we make use of below.

To use the \texttt{readMSnSet2} function, as a minimum one must
specify the file path to the data and which columns of the spreadsheet
contain quantitation data. In the code chunk below, we start by
identifying the file that we want to use. The \texttt{system.file}
function is used to return the path to the \texttt{extdata} directory
from the \Biocexptpkg{pRolocdata} package, which is where our file of
interest resides. We then use the \texttt{dir} function to list the
content of that directory and store the path that matches the file
name of interest in the \texttt{csvfile}. Note that these two lines
are only needed here to locate a file in a package; in a more general
use case, the user would define the \texttt{csvfile} variable
containing the file name of interest directly.

@@

The feature meta-data is stored in the \texttt{fData} slot and can be
accessed by \texttt{fData(hl)}. When using \texttt{readMSnSet2}, automatically,
everything that is not defined as quantitation data by \texttt{ecol} or the
feature names by \texttt{fnames} is deposited to the \texttt{fData} slot.

@@

Below, we update the replicate specific feature variable names and
remove the shared annotation. In the first line, we update only the
feature variable names 3 to 5 (by appending a \texttt{1}) and in the
second line, we apply the \texttt{updateFvarLabels} function to update
all feature variable names (by appending a \texttt{2}).

@@

<<combine, cache = cache>>=
combined <- MSnbase::combine(hyperLOPIT2015ms3r1, hyperLOPIT2015ms3r2)
combined
@

@@

In our particular case, missing values are indicative of protein
groups that were not acquired in both replicates (Figure
\ref{fig:namap}).

@@

Data quality is routinely examined through visualisation to verify
that sub-cellular niches have been separated along the gradient. Based
on De Duve's principle \cite{DeDuve:1981} proteins that co-localise in
a cell, exhibit similar quantitation profiles across the gradient
fractions employed. One approach that has been widely used to
visualise and inspect high throughput mass spectrometry-based
proteomics data is principal components analysis (PCA). PCA is one of
many dimensionality reduction methods, that allows one to effectively
summarise multi-dimensional data in to 2 or 3 dimensions to enable
visualisation. Very generally, the original continuous
multi-dimensional data is transformed into a set of orthogonal
components ordered according to the amount of variability that they
describe. The \texttt{plot2D} and \texttt{plot3D} methods in
\Biocpkg{pRoloc} allows one to plot the principal components (PCs) of
a dataset against one another, by default the first two components are
plotted on the x- and y-axis for the \texttt{plot2D} function, and
first three components are loaded for the \texttt{plot3D} function,
respectively (the \texttt{dims} argument can be used to plot other
PCs). If distinct clusters are observed, we assume that there is
organellar separation present in the data. Although, representing the
multi-dimensional data along a limited set of PCs does not give us a
hard quantitative measure of separation, it is extremely useful
summarising complex experimental information in one figure, to get a
simplified overview of the data.

@@

It is also useful to visualise the relative intensities along the
gradient to identify channels displaying particularly low yield. This
can be done using the \texttt{plotDist} and \texttt{boxplot}
functions, that plot the protein profiles occupancy along the gradient
(we also display the mean channel intensities) and a \texttt{boxplot}
of the column intensities.  In the two plots displayed on figure
\ref{fig:qcbx}, we re-order the TMT channels to pair corresponding
channels in the two replicates (rather than ordering the channels by
replicate).

@@

In the context of spatial proteomics, a marker protein is defined as a
well-known resident of a specific sub-cellular niche in a species
\textit{and} condition of interest. Applying this to machine learning
(ML), and specifically supervised learning, for the task of protein
localisation prediction, these markers constitute the labelled
training data to use as input to a classification analyses. Defining
well-known residents, and obtaining labelled training data for ML
analyses can be time consuming, but it is important to define markers
that are representative of the multivariate data space and on which a
classifier will be trained and generated. \Biocpkg{pRoloc} provides a
convenience function, \texttt{addMarkers}, to directly add markers to
an \texttt{MSnSet} object, as demonstrated in the code chunk
below. These marker sets can be accessed using the
\texttt{pRolocmarkers()} function. Marker sets are stored as a simple
named vector in R, and originate from in-house user-defined sets of
markers or from previous published studies \cite{Gatto:2014b}, which
are continuosly updated and integrated.

>>>>

To make use of the full functionality of the \Biocpkg{pRoloc} software,
users need to import their data into R and prepare them as an
\texttt{MSnSet}. The \texttt{MSnSet} is a dedicated data structure for
the efficient manipulation and processing of mass spectrometry and
proteomics data in R. Figure \ref{fig:msnset} illustrates a simplified view of the
\texttt{MSnSet} structure; there exists 3 key sub-parts (termed slots)
to such a data object: (1) the \texttt{exprs} (short for
\textit{expression} data) slot for storing the quantitation data, (2)
the \texttt{fData} slot (short for \textit{feature}-metadata) for
storing the feature meta-data, and finally (3) the \texttt{pData} slot
(short for \textit{pheno}-metadata, i.e. sample phenotypic data) for
storing the sample meta-data.

@@

Another slot of interest is \texttt{processingData}, that logs the
processing \texttt{MSnSet} objects undergo. The processing log can be
accessed with the \texttt{processingData} function and is displayed
under \textit{Processing information} in the textual object summary
when an \texttt{MSnSet}'s name is typed in the R console (see example
below).

@@

There are a number of ways to import quantitation data and create an
\texttt{MSnSet} instance. All methods are described in the
\Biocpkg{MSnbase}
\href{http://bioconductor.org/packages/release/bioc/vignettes/MSnbase/inst/doc/MSnbase-io.pdf}{input/output
  capabilities vignette}. The simplest method is to use the
function \texttt{readMSnSet2}. The function takes a single spreadsheet
file name as input and extracts the columns containing the
quantitation data, as identified by the argument \texttt{ecol}, to
create the expression data, while the other columns in the spreadsheet
are appended to the feature meta-data slot.  By example, in the code
chunk below we read in the \texttt{csv} spreadsheet containing the
quantitation data from the intersect of replicates 1 and 2 of the
mouse map \cite{hyper}, using the \texttt{readMSnSet2} function. The
data is as available online with the manuscript (see tab 2 of the
\texttt{xlsx} supplementary data set 1 in \cite{hyper}, which should
be exported as a text-based spreadsheet). It is also available as a
\texttt{csv} in the Bioconductor \Biocexptpkg{pRolocdata} data
package, which we make use of below.

To use the \texttt{readMSnSet2} function, as a minimum one must
specify the file path to the data and which columns of the spreadsheet
contain quantitation data. In the code chunk below, we start by
identifying the file that we want to use. The \texttt{system.file}
function is used to return the path to the \texttt{extdata} directory
from the \Biocexptpkg{pRolocdata} package, which is where our file of
interest resides. We then use the \texttt{dir} function to list the
content of that directory and store the path that matches the file
name of interest in the \texttt{csvfile} variable. Note that these two lines
are only needed here to locate a file in a package; in a more general
use case, the user would define the \texttt{csvfile} variable
containing the file name of interest directly.

@@

The feature meta-data is stored in the \texttt{fData} slot and can be
accessed by \texttt{fData(hl)}. When using \texttt{readMSnSet2}
everything that is not defined as quantitation data by \texttt{ecol}
is deposited to the \texttt{fData} slot.

@@

Below, we update the replicate specific feature variable names and
remove the shared annotation. In the first line, we update only the
feature variable names 3 to 5 (by appending a \texttt{1}) of the first
replicate and in the second line, we apply the
\texttt{updateFvarLabels} function to update all feature variable
names (by appending a \texttt{2}) of the second replicate. In lines 3
and 4, we retain the first 5 feature variables for the first replicate
and the relevant third to fourth variables for the second replicate.

@@

<<combine, cache = cache>>=
combined <- combine(hyperLOPIT2015ms3r1, hyperLOPIT2015ms3r2)
combined
@

@@

In our particular case, missing values are indicative of protein
groups that were not acquired in both replicates (Figure
\ref{fig:namap}, produced with the \texttt{image2} function).

@@

Data quality is routinely examined through visualisation to verify
that sub-cellular niches have been separated along the gradient. Based
on De Duve's principle \cite{DeDuve:1981} proteins that co-localise in
a cell, exhibit similar quantitation profiles across the gradient
fractions employed. One approach that has been widely used to
visualise and inspect high throughput mass spectrometry-based
proteomics data is principal components analysis (PCA). PCA is one of
many dimensionality reduction methods, that allows one to effectively
summarise multi-dimensional data in to 2 or 3 dimensions to enable
visualisation. Very generally, the original continuous
multi-dimensional data is transformed into a set of orthogonal
components ordered according to the amount of variability that they
describe. The \texttt{plot2D} and \texttt{plot3D} functions in
\Biocpkg{pRoloc} allows one to plot the principal components (PCs) of
a dataset against one another. By default, the first two components are
plotted on the x- and y-axis for the \texttt{plot2D} function, and
first three components are loaded for the \texttt{plot3D} function,
respectively (the \texttt{dims} argument can be used to plot other
PCs). If distinct clusters are observed, we assume that there is
organellar separation present in the data. Although, representing the
multi-dimensional data along a limited set of PCs does not give us a
hard quantitative measure of separation, it is extremely useful
summarising complex experimental information in one figure, to get a
simplified overview of the data.

@@

It is also useful to visualise the relative intensities along the
gradient to identify channels displaying particularly low yield. This
can be done using the \texttt{plotDist} and \texttt{boxplot}
functions, that plot the protein profiles occupancy along the gradient
(we also display the mean channel intensities below) and a
\texttt{boxplot} of the column intensities.  In the two plots
displayed on figure \ref{fig:qcbx}, we re-order the TMT channels to
pair corresponding channels in the two replicates (rather than
ordering the channels by replicate).

@@

In the context of spatial proteomics, a marker protein is defined as a
well-known resident of a specific sub-cellular niche in a species
\textit{and} condition of interest. Applying this to machine learning
(ML), and specifically supervised learning for the task of protein
localisation prediction, these markers constitute the labelled
training data to use as input to a classification analyses. Defining
well-known residents, and obtaining labelled training data for ML
analyses can be time consuming, but it is important to define markers
that are representative of the multivariate data space and on which a
classifier will be trained and generated. \Biocpkg{pRoloc} provides a
convenience function, \texttt{addMarkers}, to directly add markers to
an \texttt{MSnSet} object, as demonstrated in the code chunk
below. These marker sets can be accessed using the
\texttt{pRolocmarkers()} function. Marker sets are stored as a simple
named vector in R, and originate from in-house user-defined sets of
markers or from previous published studies \cite{Gatto:2014b}, which
are continuosly updated and integrated.
